First steps towards a formalization of Forcing

### 2018-06-24 LSFA/isabelle-zf.tex
compatibilizo relativizacion con la intro

We are not discussing how, if possible at all, to express that some
set satisfies some axiom. For example, we might well be interested in
proving $\emptyset \models \mathit{foundation}$, but there is no
sensible way to state this, without encoding formulas and axioms in
ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
  method to encode recursive definitions. In our development, although
  not reported in this paper, we use this internalization.}

\fbox{ You need this in order to formalize $\models$ as a predicate of
  type $i\ent i \ent o$. (Don't know how to put that here.)
}

 In
Isabelle/ZF there are, however, relativized versions of the
axioms. This alternative statement of the axioms are relativized with
respect to classes, which in Isabelle/ZF corresponds to predicates over
sets. For example, the relativized axiom of foundation is:
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ foundation{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isacharequal}{\isachargreater}o{\isacharparenright}\ {\isacharequal}{\isachargreater}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}M{\isacharparenright}\ {\isacharequal}{\isacharequal}\isanewline
\ \ \ \ \ \ \ \ {\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x\ {\isacharampersand}\ {\isachartilde}{\isacharparenleft}{\isasymexists}z{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ z{\isasymin}x\ {\isacharampersand}\ z\ {\isasymin}\ y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

% ∀x[c]
\noindent The relativized quantifier
\isa{{\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\
  P(x)} is a shorthand for \isa{{\isasymforall}x{\isachardot}\ M(x)
  {\isasymlongrightarrow} P(x)}. We are one step further to be able to
express that the empty set satisfies foundation; but we must turn the
set \isa{0 :: i} into a predicate. Isabelle/ZF provides the coercion
\isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i ={\isachargreater} o)}, thus we can write and prove:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ emp{\isacharunderscore}foundation\ {\isacharcolon}\ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

>>>>

%% We are not discussing how, if possible at all, to express that some
%% set satisfies some axiom. For example, we might well be interested in
%% proving $\emptyset \models \mathit{foundation}$, but there is no
%% sensible way to state this, without encoding formulas and axioms in
%% ZFC.\footnote{Indeed Paulson has encoded formulas using a very general
%%   method to encode recursive definitions. In our development, although
%%   not reported in this paper, we use this internalization.}
%% 
%% \fbox{ You need this in order to formalize $\models$ as a predicate of
%%   type $i\ent i \ent o$. (Don't know how to put that here.)
%% }

%% In
%% Isabelle/ZF there are, however, relativized versions of the
%% axioms. This alternative statement of the axioms are relativized with
%% respect to classes, which in Isabelle/ZF corresponds to predicates over
%% sets
We now discuss relativization in Isabelle/ZF. Relativized versions of the
 axioms can be found in the formalization of constructibility. For
 example, the relativized Axiom of Foundation is
\begin{isabelle}
\isacommand{definition}\isamarkupfalse%
\ foundation{\isacharunderscore}ax\ {\isacharcolon}{\isacharcolon}\ {\isachardoublequoteopen}{\isacharparenleft}i{\isacharequal}{\isachargreater}o{\isacharparenright}\ {\isacharequal}{\isachargreater}\ o{\isachardoublequoteclose}\ \isakeyword{where}\isanewline
\ \ \ \ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}M{\isacharparenright}\ {\isacharequal}{\isacharequal}\isanewline
\ \ \ \ \ \ \ \ {\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x{\isacharparenright}\ {\isasymlongrightarrow}\ {\isacharparenleft}{\isasymexists}y{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ y{\isasymin}x\ {\isacharampersand}\ {\isachartilde}{\isacharparenleft}{\isasymexists}z{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\ z{\isasymin}x\ {\isacharampersand}\ z\ {\isasymin}\ y{\isacharparenright}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}

% ∀x[c]
\noindent The relativized quantifier
\isa{{\isasymforall}x{\isacharbrackleft}M{\isacharbrackright}{\isachardot}\
  P(x)} is a shorthand for \isa{{\isasymforall}x{\isachardot}\ M(x)
  {\isasymlongrightarrow} P(x)}. In order to express that a (set)
satisfies this axiom we use  the coercion
\isa{{\isacharhash}{\isacharhash} :: i ={\isachargreater} (i
  ={\isachargreater} o)} provided by Isabelle/ZF. As a trivial example we
can show that the empty set satisfies Foundation:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ emp{\isacharunderscore}foundation\ {\isacharcolon}\ {\isachardoublequoteopen}foundation{\isacharunderscore}ax{\isacharparenleft}{\isacharhash}{\isacharhash}{\isadigit{0}}{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
